Relative parabolic factorizations of Chevalley groups

### 2015-07-29 paper.tex
Small fixes in the introduction

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item Gauss decomposition i.\,e. decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item Bass---Kolster decomposition --- $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$ is the maximal parabolic subgroup of $\G$ corresponding to $i$-th vertex of the Dynkin diagram of $\Phi$ and 
      $\U_i$, $\U_i^-$ are the unipotent radicals associated with it;
\item Dennis---Vaserstein decomposition --- $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over rings having small dimension.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, 
 while Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of a milder ring-theoretic assumption such as stable range condition.

Another advantage of parabolic factorizations is that they prove useful in the study of asymptotical properties of Chevalley groups.
The following theorem is the first such application. 
\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that
\begin{itemize} \item either $\sr(I) = 1$;
 \item or $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$. \end{itemize}
In the latter case assume additionally that Generalized Riemann Hypothesis holds.
Then for any $\Phi$ the width of the congruence subgroup $\G(\Phi, R, I)$ with respect to $z_\alpha$ does not exceed $C_1|\Phi| + C_2$. \end{thm}
\noindent In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng}.

@@

Another applications of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was settled by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

>>>>

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative versions of the following three kinds of decompositions:
\begin{itemize}
\item Gauss decomposition i.\,e. decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item Bass---Kolster decomposition $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$ is the maximal parabolic subgroup of $\G$ corresponding to $i$-th vertex of the Dynkin diagram of $\Phi$ and 
      $\U_i$, $\U_i^-$ are the unipotent radicals associated with it;
\item Dennis---Vaserstein decomposition $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

There are two main reasons to study parabolic factorizations.
Recall that a number of various factorizations of $\G$ are known over rings having small dimension.
For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, 
 while Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
The advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions such as the stable range condition.

Another advantage of parabolic factorizations is that they prove useful in the study of asymptotical properties of Chevalley groups.
The following theorem is the first such application. 
\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that
\begin{itemize} \item either $\sr(I) = 1$;
 \item or $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$. \end{itemize}
In the latter case assume additionally that Generalized Riemann Hypothesis holds.
Then for any $\Phi$ the width of the congruence subgroup $\G(\Phi, R, I)$ with respect to $z_\alpha$ does not exceed $C_1|\Phi| + C_2$. \end{thm}
\noindent In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).

@@

Another applications of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.
